---
title: "NutriPLasmaPredict_Proj"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages
```{r, echo = FALSE}
library(dplyr)
library(MASS)
library(psych)
library(broom)
```

## Import Data
```{r}
#import data
Plasma <- read.table("Plasma.txt", header = TRUE)

# Generate indices for train and test data
set.seed(23)
n = nrow(Plasma)
train_index = sample.int(n, size = 0.8*n)
train = Plasma[train_index, ]
valid = Plasma[-train_index, ]
```

## 1). Exploratory Data Analysis
### Check variable types
```{r}
# check variable type
cplasm <- sapply(Plasma, class); cplasm
sumplasm <- sapply(Plasma, summary)

# transform categorical var into factor type
Plasma$SEX <- as.factor(Plasma$SEX)
Plasma$SMOKSTAT <- as.factor(Plasma$SMOKSTAT)
Plasma$VITUSE <- as.factor(Plasma$VITUSE)
```

### Inspect multicollinearity
```{r, results="hide"}
num = subset(train, select = -c(SEX, SMOKSTAT, VITUSE) ) 
cor_num <- cor(num)
pairs(num)
```



### Inspect Categorical Variable
```{r}
#check categorical var
table(Plasma$SEX)
table(Plasma$SMOKSTAT)
table(Plasma$VITUSE)
```

```{r, results="hide"}
# check pie chart of categorical variables
plasma_cat = train[,(names(train)%in%c("SEX","SMOKSTAT", "VITUSE"))]
par(mfrow=c(1,3))
for (j in 1:3){
  n <- nrow(plasma_cat[,j])
  lbls <- names(table(plasma_cat[,j]))
  lab <- paste(lbls)
  pie(table(plasma_cat[,j]), labels = lab, main = paste("Pie chart of",colnames(plasma_cat[j])))
}
```



```{r, results="hide"}
# checking boxplot of categorical var
par(mfrow=c(1,2), mar=c(3,2,2,.5), mgp=c(3,1,0))
boxplot(Plasma$BETAPLASMA~Plasma$SEX,main='Betaplasma: side-by-side box plot by sex',xlab='sex',ylab='betaplasma', col = c("lightgrey","lightgreen"))
boxplot(Plasma$RETPLASMA~Plasma$SEX,main='Retplasma: side-by-side box plot by sex',xlab='sex',ylab='Retplasma', col =c("lightgrey","lightgreen"))

boxplot(Plasma$BETAPLASMA~Plasma$SMOKSTAT,main='Betaplasma: side-by-side box plot by smoking activities',
xlab='sex',ylab='betaplasma', col = c("lightblue", "lightpink","lightyellow"))
boxplot(Plasma$RETPLASMA~Plasma$SMOKSTAT,main='Retplasma: side-by-side box plot by smoking activities',
xlab='sex',ylab='Retplasma', col = c("lightblue", "lightpink","lightyellow"))

boxplot(Plasma$BETAPLASMA~Plasma$VITUSE,main='Betaplasma: side-by-side box plot by vitamin use',
xlab='sex',ylab='betaplasma', col = c("lightseagreen", "plum","khaki"))
boxplot(Plasma$RETPLASMA~Plasma$VITUSE,main='Retplasma: side-by-side box plot by vitamin use',
xlab='sex',ylab='Retplasma', col = c("lightseagreen", "plum","khaki"))
```



```{r, results="hide"}
# Check if there is significant differences between groups
# Use Tukey's HSD- honest significant difference
T1 <- TukeyHSD(aov(BETAPLASMA ~ SEX, Plasma))
T2 <- TukeyHSD(aov(BETAPLASMA ~ SMOKSTAT, Plasma))
T3 <- TukeyHSD(aov(BETAPLASMA ~ VITUSE, Plasma))

T4 <- TukeyHSD(aov(RETPLASMA ~ SEX, Plasma))
T5 <- TukeyHSD(aov(RETPLASMA ~ SMOKSTAT, Plasma))
T6 <- TukeyHSD(aov(RETPLASMA ~ VITUSE, Plasma))

par(mfrow=c(2,3))
plot(T1)
plot(T2)
plot(T3)
plot(T4)
plot(T5)
plot(T6)
```

### Check Numerical Variable
```{r, results="hide"}
# check numeric var
sumplasm <- sapply(num, summary)
describe(num,fast=TRUE)
knitr::kable(round(describe(num,fast=TRUE),3), format = "html")
```



```{r, results="hide"}
# check hist plot of numeric variables
par(mfrow=c(3,4))
for (i in 1:11){
  hist(num[,i], main = paste("Histogram of",colnames(num[i])), xlab = paste(colnames(num[i])))
}
```


## 2. Model Fitting
### A. Betaplasma
#### First order model
```{r, results="hide", fig.keep = "none"}
# Use stepwise regression to find first order model
beta = subset(train, select = -c(RETPLASMA) ) 
fit.0=lm(BETAPLASMA~1, data=beta) ##initial model, only intercept
step.aic=stepAIC(fit.0, scope=list(upper=~AGE+SEX+SMOKSTAT+QUETELET+VITUSE+CALORIES+FAT+FIBER+ALCOHOL+CHOLESTEROL+BETADIET+RETDIET, lower= ~1), direction="both", k=2, trace=FALSE)
step.aic$anova
```

Result : BETAPLASMA ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE

```{r, results="hide"}
# Diagnostic of BETA model1 
train_pos = beta[beta$BETAPLASMA > 0, ]
betamodel1 <- lm(BETAPLASMA ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE, data=train_pos)
par(mfrow = c(1,3), mar=c(3,2,2,.5), mgp=c(3,1,0))
plot(betamodel1, which=c(1,2))
MASS::boxcox(betamodel1)
```



```{r}
# log-transformation
log_betamodel1 <- lm(log(BETAPLASMA) ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE, data=train_pos)
summary(log_betamodel1)
```

```{r, results="hide"}
# diagnosis of log
par(mfrow = c(1,3), mar=c(3,2,2,.5), mgp=c(3,1,0))
plot(log_betamodel1, which= c(1,2))
MASS::boxcox(log_betamodel1)
# hist plot
par(mfrow = c(1,2), mar=c(3,2,2,.5), mgp=c(3,1,0))
hist(train$BETAPLASMA, main = paste("Histogram of betaplasma", xlab = "betaplasma"))
hist(log(train$BETAPLASMA), main = paste("Histogram of log betaplasma", xlab = "log betaplasma"))
par(mfrow=c(1,1), mar=c(3,2,2,.5), mgp=c(3,1,0)) 
# linearity check
num$BETADIET <- log(num$BETADIET)
pairs(num)
```



#### Model with interactive terms
```{r, results="hide"}
#Adding interaction effects
#betamodel1 = lm(BETAPLASMA ~ FIBER + QUETELET + VITUSE + FAT + BETADIET, data = beta)
stepbeta <- step(betamodel1, scope = FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE ~ .^2, direction = 'both')
```

Result : BETAPLASMA ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE + 
    FIBER:VITUSE + FIBER:BETADIET + FAT:BETADIET + BETADIET:AGE + 
    FAT:AGE

```{r, results="hide"}
# Diagnostic of BETA model2
betamodel2 <- lm(formula = BETAPLASMA ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE + 
          FIBER:VITUSE + FIBER:BETADIET + FAT:BETADIET + BETADIET:AGE + FAT:AGE, data = train_pos)

par(mfrow = c(1,3))
plot(betamodel2, which=c(1,2))
MASS::boxcox(betamodel2)
```



```{r}
# log-transformation
log_betamodel2 <- lm(formula = log(BETAPLASMA) ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE + FIBER:VITUSE + FIBER:BETADIET + FAT:BETADIET + BETADIET:AGE + FAT:AGE, data = train_pos)
summary(log_betamodel2)
```

```{r, results="hide"}
# diagnosis of log
par(mfrow = c(1,3), mar=c(3,2,2,.5), mgp=c(3,1,0))
plot(log_betamodel2, which=c(1,2))
MASS::boxcox(log_betamodel2)
summary(log_betamodel2)
```



### B. Retoplasma
#### First order model
```{r, results="hide"}
# Stepwise Regression for first order
ret = subset(train, select = -c(BETAPLASMA) ) 
fit.02=lm(RETPLASMA~1, data=ret) ##initial model, only intercept
step.aic2=stepAIC(fit.02, scope=list(upper=~AGE+SEX+SMOKSTAT+QUETELET+VITUSE+CALORIES+FAT+FIBER+ALCOHOL+CHOLESTEROL+BETADIET+RETDIET, lower= ~1), direction="both", k=2, trace=FALSE)
step.aic2$anova
```

Result : RETPLASMA ~ SEX + CALORIES + AGE

```{r, results="hide"}
# Diagonistic of RET model1
retmodel1 <- lm(formula = RETPLASMA ~ SEX + CALORIES + AGE, data = ret)
par(mfrow = c(1,3))
plot(retmodel1, which = c(1,2))
MASS::boxcox(retmodel1)
```



```{r}
# log-transformation
log_retmodel1 <- lm(formula = log(RETPLASMA) ~ SEX + CALORIES + AGE, data = ret)
summary(log_retmodel1)
```

```{r, results="hide"}
# diagnosis of log
par(mfrow = c(1,3))
plot(log_retmodel1, which=c(1,2))
MASS::boxcox(log_retmodel1)
summary(log_retmodel1)
# hist plot
par(mfrow = c(1,2), mar=c(3,2,2,.5), mgp=c(3,1,0))
hist(train$RETPLASMA, main = paste("Histogram of retplasma", xlab = "retplasma"))
hist(log(train$RETPLASMA), main = paste("Histogram of log retplasma", xlab = "log retplasma"))
par(mfrow=c(1,1), mar=c(3,2,2,.5), mgp=c(3,1,0)) 
# linearity check
num$RETPLASMA <- log(num$RETPLASMA)
pairs(num)
```



#### Model with interactive terms
```{r}
#Adding interactive terms
#init_mod2 <- lm(RETPLASMA ~ SEX + CALORIES + AGE, data = ret)
stepret <- step(retmodel1, scope = SEX + CALORIES + AGE ~ .^2, direction = 'both')
```

Result : RETPLASMA ~ SEX + CALORIES + AGE + SEX:AGE + SEX:CALORIES

```{r, results="hide"}
# Diagonistic of RET model2
retmodel2 <- lm(formula = RETPLASMA ~ SEX + CALORIES + AGE + SEX:AGE + SEX:CALORIES, data = ret)
par(mfrow = c(1,3))
plot(retmodel1, which = c(1,2))
MASS::boxcox(retmodel1)
```



```{r}
# log-trans and diagonistic
log_retmodel2 <- lm(formula = log(RETPLASMA) ~ SEX + CALORIES + AGE + SEX:AGE + SEX:CALORIES, data = ret)
summary(log_retmodel2)
```

```{r, results="hide"}
par(mfrow = c(1,3))
plot(log_retmodel2, which = c(1,2))
MASS::boxcox(log_retmodel2)
```



## 3. Model Selection: Criterion
```{r}
# get rsq, radj, bic, aic
sumb1 <- glance(log_betamodel1)
sumb2 <- glance(log_betamodel2)
sumr1 <- glance(log_retmodel1)
sumr2 <- glance(log_retmodel2)

# get sse
anvb1 <- anova(log_betamodel1)
anvb2 <- anova(log_betamodel2)
anvr1 <- anova(log_retmodel1)
anvr2 <- anova(log_retmodel2)
```

```{r}
# Get cp
# Full model for models with interaction
fullmod1 <- lm(log(BETAPLASMA)~.^2, data=train_pos)
fullmod2 <- lm(log(RETPLASMA)~.^2, data= ret)
anvf1 <- anova(fullmod1)
anvf2 <- anova(fullmod2)
MSEf1 <- anvf1$`Mean Sq`[79]
MSEf2 <- anvf2$`Mean Sq`[79]
```

```{r}
#Full model for first order model
fullmod11 <- lm(log(BETAPLASMA)~., data=train_pos)
fullmod22 <- lm(log(RETPLASMA)~., data=ret)
anvf11 <- anova(fullmod11)
anvf22 <- anova(fullmod22)
MSEf11 <- anvf11$`Mean Sq`[13]
MSEf22 <- anvf22$`Mean Sq`[13]
```

```{r}
# Calculate cp
cpb1 <- (anvb1$`Sum Sq`[7]/MSEf11) - (252-(2*7))
cpb2 <- (anvb2$`Sum Sq`[12]/MSEf1) - (252-(2*12))
cpr1 <- (anvr1$`Sum Sq`[4]/MSEf22) - (252-(2*4))
cpr2 <- (anvr2$`Sum Sq`[6]/MSEf2) - (252-(2*6))
```

```{r}
#Get pressp
# 1. log_betamodel1
eb1 = log_betamodel1$residuals
hb1 = influence(log_betamodel1)$hat
de_b1 = eb1/(1-hb1)
pressb1 = sum((de_b1)^2)

# 2. log_betamodel2
eb2 = log_betamodel2$residuals
hb2 = influence(log_betamodel2)$hat
de_b2 = eb2/(1-hb2)
pressb2 = sum( de_b2^2 )

# 3. log_retmodel1
er1 = log_retmodel1$residuals
hr1 = influence(log_retmodel1)$hat
de_r1 = er1/(1-hr1)
pressr1 = sum( de_r1^2 )

# 4. log_retmodel2
er2 = log_retmodel2$residuals
hr2 = influence(log_retmodel2)$hat
de_r2 = er2/(1-hr2)
pressr2 = sum( de_r2^2 )
```

```{r}
mod <- c("Beta1","Beta2", "Ret1", "Ret2")
sse <- round(c(anvb1$`Sum Sq`[7], anvb2$`Sum Sq`[12], anvr1$`Sum Sq`[4], anvr2$`Sum Sq`[6]),3)
rsq <- round(c(sumb1$r.squared, sumb2$r.squared, sumr1$r.squared, sumr2$r.squared ),3)
rsqa <- round(c(sumb1$adj.r.squared, sumb1$adj.r.squared, sumr1$adj.r.squared, sumr2$adj.r.squared),3)
cp <- round(c(cpb1, cpb2, cpr1, cpr2),3)
bic <- round(c(sumb1$BIC, sumb2$BIC, sumr1$BIC, sumr2$BIC),3)
aic <- round(c(sumb1$AIC, sumb2$AIC, sumr1$AIC, sumr2$AIC),3)
pressp <- round(c(pressb1, pressb2, pressr1, pressr2),3)
res_sub <- cbind(mod, sse, rsq, rsqa, cp, bic, aic, pressp)
colnames(res_sub)<-c("Model", "sse", "R^2", "R^2_a", "Cp","bic", "aic", "press_p")
as_tibble(res_sub)
```

## 4. Data Validation
```{r}
# Data Validation (Beta)
beta_train <- log_betamodel1
beta_valid <- lm(formula = log(BETAPLASMA) ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE, data = valid)

mod_sum <- cbind(coef(summary(beta_train))[,1], coef(summary(beta_valid))[,1],
coef(summary(beta_train))[,2], coef(summary(beta_valid))[,2])
colnames(mod_sum) <- c("Train Est","Valid Est","Train s.e.","Valid s.e.")

mod_sum
```

```{r}
#compare the SSE and R2a (Beta)
sse_t <- sum(beta_train$residuals^2)
n_t = nrow(train)
mse_t <- sse_t/(n_t-7)
sse_v <- sum(beta_valid$residuals^2)
n_v = nrow(valid)
mse_v <- sse_v/(n_v-7)
Radj_t <- summary(beta_train)$adj.r.squared
Radj_v <- summary(beta_valid)$adj.r.squared
train_sum <- c(sse_t,mse_t,Radj_t)
valid_sum <- c(sse_v,mse_v,Radj_v)
criteria <- rbind(train_sum,valid_sum)
colnames(criteria) <- c("SSE","MSE","R2_adj")
criteria
```

```{r}
#Get MSPE_v from new data (beta)
beta_newdata <- data.frame(valid[, 1:12])
y.hat <- predict(beta_train, beta_newdata)
MSPE <- mean((log(valid$BETAPLASMA) - y.hat)^2)
criteria <- cbind(MSPE,sse_t/n_t)
colnames(criteria) <- c("MSPE","SSE/n")
criteria
```

```{r}
# Data Validation (Ret)
ret_train <- log_retmodel1
ret_valid <- lm(formula = log(RETPLASMA) ~ SEX + CALORIES + AGE, data = valid)

mod_sum <- cbind(coef(summary(ret_train))[,1], coef(summary(ret_valid))[,1],
coef(summary(ret_train))[,2], coef(summary(ret_valid))[,2])
colnames(mod_sum) <- c("Train Est","Valid Est","Train s.e.","Valid s.e.")

mod_sum
```

```{r}
#compare the SSE and R2a (Ret)
sse_t <- sum(ret_train$residuals^2)
n_t = nrow(train)
mse_t <- sse_t/(n_t-7)
sse_v <- sum(ret_valid$residuals^2)
n_v = nrow(valid)
mse_v <- sse_v/(n_v-7)
Radj_t <- summary(ret_train)$adj.r.squared
Radj_v <- summary(ret_valid)$adj.r.squared
train_sum <- c(sse_t,mse_t,Radj_t)
valid_sum <- c(sse_v,mse_v,Radj_v)
criteria <- rbind(train_sum,valid_sum)
colnames(criteria) <- c("SSE","MSE","R2_adj")
criteria
```

```{r}
#Get MSPE_v from new data (ret)
ret_newdata <- data.frame(valid[, 1:12])
y.hat <- predict(ret_train, ret_newdata)
MSPE <- mean((log(valid$RETPLASMA) - y.hat)^2)
criteria <- cbind(MSPE,sse_t/n_t)
colnames(criteria) <- c("MSPE","SSE/n")
criteria
```

## 5. Rebuild model use whole dataset
```{r}
Plasma_pos = Plasma[Plasma$BETAPLASMA > 0, ]
log_betamodel_final <- lm(log(BETAPLASMA) ~ FIBER + QUETELET + VITUSE + FAT + BETADIET + AGE, data = Plasma_pos)
summary(log_betamodel_final)
log_retmodel_final <- lm(formula = log(RETPLASMA) ~ SEX + CALORIES + AGE, data = Plasma)
summary(log_retmodel_final)
```

